---
title: "Reproducible Research, Project 1"
output: md_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

The purpose of this report is to perform an analysis on a sample data set from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

##Setup
Set the working directory:
```{r, echo = TRUE}
directory <- "C:/Users/Public"
setwd(directory)
```
Load the necessary packages:
```{r, echo = TRUE, message = FALSE}
library(ggplot2)
library(dplyr)
library(chron)
```
Download file, unzip, and read file into R, and process the data. 
```{r, echo = TRUE, warning = FALSE}
zip_file <- download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
              destfile = "C:/Users/Public/data.zip")
unzip("data.zip")
activity <- read.csv("activity.csv")
activity$Day <- weekdays(as.Date(activity$date))
activity_no_NA <- activity[!is.na(activity$steps),]
```
## Part 1: What is the mean total number of steps taken per day?
Calculate the total number of steps taken for each day in the dataset. 
```{r, echo = TRUE}
sum_steps <- activity_no_NA %>%
  group_by(date) %>%
  summarize(steps = sum(steps, na.rm = TRUE))
```
A histogram of the total number of steps taken each day can now be created:
```{r, echo = TRUE}
p <- ggplot(sum_steps, aes(x=steps)) + 
  geom_histogram(binwidth = 1000, color="white", fill="navy") +
  labs(title="Steps Taken per Day",x="Average Steps", y = "Count")
p
```
```{r, echo = TRUE}
mean_num_steps <- as.integer(mean(sum_steps$steps))
```
The mean of the total number of steps taken per day is: `r mean_num_steps`

```{r, echo = TRUE}
median_num_steps <- as.integer(median(sum_steps$steps))
```
The median of the total number of steps taken per day is: `r median_num_steps`

## Part 2: What is the average daily activity pattern?
Calculate the mean number of steps, grouped by each 5-minute interval.
```{r, echo = TRUE}
Int <- activity_no_NA %>%
  group_by(interval) %>%
  summarize(Avg = mean(steps))
```
A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, 
averaged across all days (y-axis) can now be created:
```{r, echo = TRUE}
q <- ggplot(Int, aes(x=interval, y=Avg)) +
  geom_line(size = 1, color = "navy") +
  labs(title = "Average Steps per Interval", x = "Interval", y = "Steps Average")
q
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}
summarize(Int, as.integer(max(Avg)))
Int[Int$Avg==max(Int$Avg),1]
```
## Part 3: Imputing Missing Values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
```{r, echo = TRUE}
sum_na_activity <- sum(is.na(activity$steps))
```
The total number of missing values in the dataset is: `r sum_na_activity`

The strategy used for filling in missing step values in the dataset is using the mean for that 5-minute interval.
```{r, echo = TRUE}
avg_steps_per_int <- activity_no_NA %>%
  group_by(interval) %>%
  summarize(steps = as.integer(mean(steps)))
```
Combine missing steps data with the average steps per 5-min interval data.
```{r, echo = TRUE}
na_activity <- activity[is.na(activity$steps),]
imp_activity <- merge(na_activity, avg_steps_per_int, by = c("interval"))
imp_activity <- rename(imp_activity, steps = steps.y)
imp_activity$steps.x <- NULL
imp_activity_RE <- imp_activity[,c(4,2,1,3)]
```
Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo= TRUE}
merge_imp <- bind_rows(activity_no_NA, imp_activity_RE)
```
Calcuate the average number of steps taken per day.
```{r, echo = TRUE}
sum_steps_imp <- merge_imp %>%
  group_by(date) %>%
  summarize(steps = sum(steps))
```
A histogram of the total number of steps taken each day for the imputed dataset can now be created:
```{r, echo = TRUE}
r <- ggplot(sum_steps_imp, aes(x=steps)) + 
  geom_histogram(binwidth = 1000, color="white", fill="navy") +
  labs(title="Steps Taken per Day (Imputed via 5 minute Interval Averages)",x="Steps", y = "Count")
r
```
```{r, echo = TRUE}
mean_num_steps_imp <- as.integer(mean(sum_steps_imp$steps))
```
The mean of the total number of steps taken per day for the imputed data set is: `r mean_num_steps_imp`

```{r, echo = TRUE}
median_num_steps_imp <- as.integer(median(sum_steps_imp$steps))
```
The median of the total number of steps taken per day for the imputed data set is: `r median_num_steps_imp`

The values of the imputed data set (mean, median = `r mean_num_steps_imp`, `r median_num_steps_imp`) are different
from the original data set (mean, median =  `r mean_num_steps`, `r median_num_steps`). The mean and median of the imputed dataset is lower than that original dataset, which is reasonable since the imputed data strategy involved using the averages of the 5-minute intervals for the missing data. 

## Part 4: Are there differences in activity patterns between weekdays and weekends?
Identify if a day is a weekend (TRUE) or a weekday (FALSE):
```{r, echo = TRUE}
merge_imp$is_Weekend <- is.weekend(merge_imp$date)
merge_imp$Weekend <- ifelse (merge_imp$is_Weekend == TRUE, "Weekend", "Weekday")
```
Calculate the average number of steps for each interval, grouped by weekends and weekdays. 
```{r, echo = TRUE}
Int_wk <- merge_imp %>%
  group_by(interval,Weekend) %>%
  summarize(Avg = mean(steps))
```
A panel plot containing a time series plot of the 5-minute interval (x-axis) and the average 
number of steps taken, averaged across all weekday days or weekend days (y-axis) can now be created:
```{r, echo = TRUE}
s <- ggplot(Int_wk, aes(x=interval, y=Avg)) +
  facet_grid(Weekend ~.) +
  geom_line(size = 1, color = "navy") +
  labs(title = "Average Steps per Interval; Weekdays vs. Weekends", x = "Interval", y = "Steps Average")
s
```

